Improving Haskell Types with SMT

### 2015-07-18 docs/paper.tex
Improvements based on reviewer feedback (and fix page breaks)

Next, we define a function to add two such numbers:
\begin{Verbatim}
uAdd :: UNat m -> UNat n -> UNat (m + n)
uAdd Zero     y  = y
uAdd (Succ x) y  = Succ (uAdd x y)
\end{Verbatim}
While this is a simple definition, and we are unlikely
to have gotten it wrong, it is nice to know that GHC
is checking our work!  Had we made a mistake, for
example, by mis-typing the recursive call as \Verb"uAdd x x",
we would get a type error:
\begin{Verbatim}
  Could not deduce (((n1 + n1) + 1) ~ (m + n))
    from the context (m ~ (n1 + 1))
\end{Verbatim}
It is a small detail, but it is worth pointing out that
GHC would have been just as happy had we defined the type
of \Verb"Succ" like this:
\begin{Verbatim}
  Succ :: UNat n -> UNat (1 + n)
\end{Verbatim}
The difference is in the return type, in the one case it is
\Verb"n + 1", and in the other it is \Verb"1 + n".  While
to a human this looks like an insignificant difference,
in many systems it is significantly easier to work
with the one definition, but not the other.

@@

While these examples are somewhat simplistic, we hope
that they demonstrate the utility of the technology,
and show that using the extension feels quite natural
to a Haskell programmer.

@@

@@

mustBeK :: Solver -> Name -> Value -> IO Bool
mustBeK s x v =
  solverProve s (eq (SMT.const x) (value v))
\end{Verbatim}
This process takes $O(n)$ time, where $n$ is the number of variables.
Note that here we are making use of the incremental capabilities of the
solver, and reusing all asserted constraints, just doing one
additional \Verb"assert" per variable. To see the process in action,
consider the constraint \Verb"(2 * x) ~ 16".  The steps that we'd perform
are:
\begin{Verbatim}
(assert (= (* 2 x) 16))
(check-sat)
; SMT response: SAT
(get-value (x))
; SMT response: (8)
(push 1)
(assert (not (= x 8)))
(check-sat)
; SMT response: UNSAT
(pop 1)
; We generate a computed improvement: x ~ 8
\end{Verbatim}

@@

We admit that this improvement is certainly somewhat bit ad-hoc: after all,
why look for relations between two variables and more?  We implemented
this improvement, because while experimenting with the implementation,
we noticed that this sort of improvement is needed quite often,
at least for a certain class of problems.  So, as far as we know, this
simply a useful heuristic that is worth considering.  It is possible
that alternative techniques (e.g., custom rewrite rules) could achieve
the same effect more directly, and we are certainly considering exploring
such approaches in future work.

To see when this improvement is needed, consider the following program
fragment:
\begin{Verbatim}
f :: Proxy (a + 1)
f = Proxy

g :: Proxy (b + 2)
g = f
\end{Verbatim}
In this example, GHC needs to infer how to instantiate \Verb"f"'s type
parameter, \Verb"a", and to do so, our plug-in needs to figure out that
\Verb"a" may be instantiate to \Verb"b + 1".

Once we compute candidate values for $A$ and $B$, we still need to
invoke the solver to validate that the compute relation holds in all
instances of the constraints, and not just the two that we happened to pick.

@@

\paragraph{Custom Improving Rewrites.}
All the improvements that we've defined so far work by {\em only} using
the solver---note that we were just looking at satisfying assignments,
and did not have any special rules about the shapes of the constraints.
It also makes sense to extend the system with custom rewrite rules,
which has the potential of speeding up performance, and also adding
support for features that go beyond the solver's capabilities.
We have experimented with various custom rules, but it is as yet
unclear what constitutes a good set of rewrites.

@@

The option \Verb"fplugin=TypeNatSolver" instructs GHC to load the SMT solver
plug-in.  The rest are Haskell extensions that are commonly used in this
setting:
\begin{itemize}
\item \Verb"DataKinds" enables natural number literals at the type
level.
\item \Verb"TypeOperators" and \Verb"TypeFamilies" are need to work
with type-level functions such as \Verb"+".
\item \Verb"KindSignatures" is used to specify the kinds of type variables,
for example, to state that some type-variable is of kind \Verb"Nat".
\item \Verb"GADTs" enables Generalized Algebraic Data-types;  strictly
speaking, these are not required to experiment with the plug-in, however,
many of the more interesting uses of the plug-in end up using GADTs.
\end{itemize}

>>>>

Next, we define a function to add two such numbers:
\begin{Verbatim}
uAdd :: UNat m -> UNat n -> UNat (m + n)
uAdd Zero     y  = y
uAdd (Succ x) y  = Succ (uAdd x y)
\end{Verbatim}
While this is a simple definition, and we are unlikely
to have gotten it wrong, it is nice to know that GHC
is checking our work!  Had we made a mistake, for
example, by mis-typing the recursive call as \Verb"uAdd x x",
we would get a type error:
\begin{Verbatim}
  Could not deduce (((n1 + n1) + 1) ~ (m + n))
    from the context (m ~ (n1 + 1))
\end{Verbatim}
It is a small detail, but it is worth pointing out that our
GHC extension would have been just as happy had we defined the type
of \Verb"Succ" like this:
\begin{Verbatim}
  Succ :: UNat n -> UNat (1 + n)
\end{Verbatim}
The difference is in the return type, in the one case it is
\Verb"n + 1", and in the other it is \Verb"1 + n".  While
to a human this looks like an insignificant difference,
in many systems it is significantly easier to work
with the one definition, but not the other.

@@

Note that while these examples are somewhat simplistic,
they demonstrate the utility of the technology, and show that using
the extension feels quite natural to a Haskell programmer.

A more conventional example is the type of fixed-length vectors:
\begin{Verbatim}
data Vec :: Nat -> * -> * where
  Nil  :: Vec 0 a
  Cons :: a -> Vec n a -> Vec (n + 1) a
\end{Verbatim}
Note that this is very similar to \Verb"UNat", except that it also
stores a value in each \Verb"Cons" node.  The operation of appending
two vectors together follows exactly the same pattern as \Verb"uAdd".
An example of a function that is meaningful for vectors, bit no
so much for the unary natural numbers is reversing a vector, which
may be defined as follows using our extension:
\begin{Verbatim}
vecReverse = go Nil
  where
  go :: Vec m a -> Vec n a -> Vec (m + n) a
  go xs Nil = xs
  go xs (Cons y ys) = go (Cons y xs) ys
\end{Verbatim}
This example also illustrates that our extension does not change any type
signature requirements beyond what is already in GHC: the helper
function \Verb"go" needs a type signature, because it pattern-matches
using GADT constructors, however the type of \Verb"reverse" can be
inferred as usual:
\begin{Verbatim}
*Example> :type vecReverse
vecReverse :: Vec n a -> Vec n a
\end{Verbatim}

@@

\break

@@

mustBeK :: Solver -> Name -> Value -> IO Bool
mustBeK s x v =
  solverProve s (eq (SMT.const x) (value v))
\end{Verbatim}
This process takes $O(n)$ time, where $n$ is the number of variables.
Note that here we are making use of the incremental capabilities of the
solver, and reusing all asserted constraints, just doing one
additional \Verb"assert" per variable. To see the process in action,
consider the constraint \Verb"(2 * x) ~ 16".  The steps that we'd perform
are:
\break
\begin{Verbatim}
(assert (= (* 2 x) 16))
(check-sat)
; SMT response: SAT
(get-value (x))
; SMT response: (8)
(push 1)
(assert (not (= x 8)))
(check-sat)
; SMT response: UNSAT
(pop 1)
; We generate a computed improvement: x ~ 8
\end{Verbatim}

@@

Once we compute candidate values for $A$ and $B$, we still need to
invoke the solver to validate that the computed relation holds in all
instances of the constraints, and not just the two that we happened to pick.

\break
To see when this improvement is needed, consider the following program
fragment:
\begin{Verbatim}
f :: Proxy (a + 1)
f = Proxy

g :: Proxy (b + 2)
g = f
\end{Verbatim}
To check this program, GHC needs to solve the constraint \\
\Verb"(?a + 1) ~ (b + 2)". The question mark in \Verb"?a" emphasizes that
this is a unification variable, standing-in for a type that needs to be
inferred. We cannot discharge this constraint immediately, because it
does not hold {\em for all} possible values of \Verb"?a" and \Verb"b".
However, since \Verb"?a" is a unification variable, we have the option
of instantiating it to change the constraint.  Of course, we have to
be careful that in doing so we do not loose any generality and, in this
case, the safe instantiation is \Verb"?a := b + 1".  The purpose of the
linear-relation improvement is to discover these kinds of
instantiations.

At first sight, this example may appear artificial, however it is an
abstraction of a very common pattern: namely when one function calls
another, and the type of the provided argument does not match
syntactically the type of the expected argument.  As another example,
consider the function \Verb"vecHead", which requires that its parameter
has at least one element:
\begin{Verbatim}
vecHead :: Vec (n + 1) a -> a
vecHead (Cons x _) = x
\end{Verbatim}
Now, if we tried to call \Verb"vecHead" with an expression of type
\Verb"Vec (b + 2) a", we'd end up in exactly the same situation as the
previous example, where we'd have to infer that \Verb"?n = b + 1".

@@

We admit that this improvement is somewhat ad-hoc, but as the previous
examples illustrate, it helps in situations that are quite common in
practice.  It was also fairly simple to implement using the solver.

\paragraph{Custom Improving Rewrites.}
All the improvements that we've defined so far work by {\em only} using
the solver---note that we were just looking at satisfying assignments,
and did not have any special rules about the shapes of the constraints.
It also makes sense to extend the system with custom rewrite rules,
which has the potential of speeding up performance, and also adding
support for features that go beyond the solver's capabilities.
We have experimented with various custom rules, but it is as yet
unclear what constitutes a good set of rewrites.  Concrete custom
rules that would help are various normalization rules, for example
applying distributivity repeatedly, and also rules that rewrite
equations so that there is just a unification variable on the one
side of the equation, which would help it get instantiated.

@@

The option \Verb"fplugin=TypeNatSolver" instructs GHC to load the SMT solver
plug-in.  The rest are Haskell extensions that are commonly used in this
setting:

\break
\begin{itemize}
\item \Verb"DataKinds" enables natural number literals at the type
level.
\item \Verb"TypeOperators" and \Verb"TypeFamilies" are need to work
with type-level functions such as \Verb"+".
\item \Verb"KindSignatures" is used to specify the kinds of type variables,
for example, to state that some type-variable is of kind \Verb"Nat".
\item \Verb"GADTs" enables Generalized Algebraic Data-types;  strictly
speaking, these are not required to experiment with the plug-in, however,
many of the more interesting uses of the plug-in end up using GADTs.
\end{itemize}
